Beating Simple Backtracking Search in Subgraph Algorithms\thanks{This work was supported by the Engineering and Physical Sciences Research Council [grant numbers EP/026842/1 and EP/M508056/1].}

### 2018-11-26 paper/paper.tex
More things

\begin{abstract}
    The current state of the art in subgraph isomorphism solving involves using degree as a
    value-ordering heuristic to direct backtracking search. Such a search makes a heavy commitment
    to the first branching choice, which is often incorrect. To mitigate this, we introduce and
    evaluate a new approach, which we call ``solution-biased search''. By combining a
    slightly-random value-ordering heuristic, rapid restarts, and nogood recording, we achieve two
    orders of magnitude faster performance on satisfiable instances, whilst not affecting
    performance on unsatisfiable instances. This search can be parallelised in a very simple
    but effective way: across both satisfiable and unsatisfiable instances, we get a further speedup
    of over thirty from thirty-six cores, or one hundred from ten distributed-memory hosts.  We also
    show that solution-biased search is also suitable for optimisation problems, by using it to
    improve two maximum common induced subgraph algorithms.
\end{abstract}

@@

The subgraph isomorphism problem is to decide whether a copy of a small ``pattern'' graph occurs
inside a larger ``target'' graph. Although the problem is \NP-complete, by combining design
techniques from artificial intelligence with careful algorithm engineering, modern subgraph
isomorphism solvers can often produce exact solutions quickly even on graphs with thousands of
vertices. The current single strongest subgraph isomorphism solver uses ``highest degree first'' as
a value-ordering heuristic to direct a constraint programming style search
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16,DBLP:journals/jair/McCreeshPST18}. This
heuristic is much better than branching randomly, but is still far from perfect. To offset mistakes
made by this heuristic, this paper proposes a new perspective on value-ordering: rather than
defining a search order, we use degree to direct what \emph{proportion} of the search effort should
be spent in each subproblem. By combining rapid restarts and nogood recording, and introducing a
small amount of randomness into the value-ordering heuristic, we make a state-of-the-art subgraph
algorithm perform two orders of magnitude better on a large number of satisfiable instances, whilst
performing worse only rarely on satisfiable instances, and never on unsatisfiable instances.

This new form of search can also be parallelised. By running many threads with different random
seeds but the same restart schedule, and sharing nogoods only following restarts, we can achieve
speedups of ??, with a \emph{much} simpler implementation than conventional work-stealing.  Finally,
we show that this strategy is also effective in an optimisation setting, producing benefits in two
maximum common induced subgraph algorithms. This combination of techniques gives us, for the first
time, a practical alternative to backtracking search for constraint programming style algorithms
where we have a strong value-ordering heuristic and where we care both about satisfiable and
unsatisfiable instances.

@@

?? Applications

@@

\begin{description}
    \item[BVG(r), M4D(r), and Rand] are families of randomly generated graphs using different models (bounded
        degree, regular mesh, and uniform), where each pattern is a permuted random connected subgraph
        of the target (and so each instance is satisfiable) \cite{DBLP:journals/pami/CordellaFSV04}.
        These benchmark instances are widely used, but have unusual properties and so broad
        conclusions should not be drawn based solely upon behaviour on these instances
        \cite{DBLP:journals/jair/McCreeshPST18}.
    \item[SF] contains randomly generated scale-free graphs using a similar method
        \cite{DBLP:journals/constraints/ZampelliDS10}.
   \item[LV] consists of various kinds of graph gathered by Larrosa and Valiente
       \cite{DBLP:journals/mscs/LarrosaV02} from the Stanford Graph Database. We include both the
        50 small graphs, and the 50 larger graphs.
    \item[Phase] contains hard crafted instances that lie near the satisfiable / unsatisfiable phase
        transition \cite{DBLP:journals/jair/McCreeshPST18}.
    \item[PR] contains graphs generated from segmented images, corresponding to a computer vision
        problem \cite{DBLP:journals/pr/SolnonDHJ15}.
    \item[Images and Meshes] contain graphs representing 2D segmented images and 3D object models
        respectively, again representing a computer vision problem
        \cite{DBLP:journals/cviu/DamiandSHJS11}.
\end{description}

@@

We begin with a set of baseline performance measurements. In the first and
third plots \cref{figure:others} we show the cumulative number of instances solved over time for the
non-induced and induced problems respectively. We compare the Glasgow Subgraph Solver using
depth-first search (DFS) and with the modifications described in the remainder of this paper
(solution-biased search, SBS), the PathLAD variation of the LAD algorithm
\cite{DBLP:journals/ai/Solnon10,DBLP:conf/lion/KotthoffMS16}, VF2
\cite{DBLP:journals/pami/CordellaFSV04}, RI \cite{DBLP:journals/bmcbi/BonniciGPSF13}, and VF3
\cite{DBLP:conf/gbrpr/CarlettiFSV17} (which only supports the induced problem), in each case using
the original implementation provided by the algorithm's authors. For each algorithm, the $y$ value
gives the cumulative number of instances which (individually) can be solved in no more than $x$
milliseconds.  The vertical distance between two lines therefore shows how many more instances can
be solved by one solver than another, if every instance is run separately with the chosen $x$
timeout. The horizontal distance shows how many times longer the per-instance timeout would need to
be to allow the rightmost algorithm to succeed on $y$ out of the 14,621 instances (bearing in mind
that the two sets of $y$ instances could be different).  The plots show that our starting point
comfortably beats PathLAD, VF2, VF3 and RI, except for very low choices of timeout.

@@

Traditionally, value-ordering heuristics are designed to drive search towards the most promising
region of the search space ?? cite, or the most constrained ?? cite, or the region with the highest solution
density ?? cite. In subgraph isomorphism, this is done by selecting vertices from highest degree to
lowest \cite{DBLP:journals/jair/McCreeshPST18}. The left-hand plot of
\cref{figure:value-ordering-heuristics} demonstrates that this is indeed a good choice: the
\textbf{Degree} heuristic's line shows much better performance on satisfiable instances than the
\textbf{Random} (branch randomly) or \textbf{Anti} (branch from lowest degree to highest) heuristic
lines. (On unsatisfiable instances, the value-ordering heuristic has no effect on performance,
because a complete search must be performed.)

But what happens if our value-ordering heuristic has to choose between mapping a pattern vertex to,
for example, three vertices of degree ten, two vertices of degree nine, or five of degree two? When
driving conventional backtracking search, the degree heuristic would pick one of the vertices of
degree ten, and we would commit all of our search effort to the exponentially large search tree
underneath it.  We will show that this is not a wise choice, and that instead, we should \emph{commit
equal search effort} to each of the three subproblems found by mapping to vertices of degree ten.
And similarly, should we be certain that vertex of degree ten is so much better than a vertex of
degree nine that we should commit no effort to degree nine vertices until all the degree ten
subproblems have been explored? Or might it be better to commit, say, twice as much effort to each
degree ten subproblems as to each degree nine subproblems, and only a very small amount of effort to
the degree two subproblems?  To test this hypothesis, we will now introduce a new alternative to
backtracking search, which we call \emph{solution-biased search}.  This search is made up of three
components: a new slightly-random value-ordering heuristic, rapid restarts, and nogood recording.

@@

We define a new \textbf{Biased} value-ordering heuristic, as follows. When branching, we select a
vertex $v'$ from the chosen domain $D_v$ with probability \[ p(v') = \frac{2^{\deg(v')}}{\sum_{w \in
D_v}{2^{\deg(w)}}} \text{.} \] This heuristic is now equally likely to pick between vertices of
equal degree, is twice as likely to pick a vertex of degree $d$ as one of degree $d - 1$, and is
over a thousand times more likely to pick a vertex of degree $d$ than degree $d - 10$.
\Cref{figure:value-ordering-heuristics} shows that the performance of the degree and biased
heuristics are very similar: the biased heuristic has slightly worse performance below a ??
timeout, and effectively equal performance above it. This shows that we can introduce an element of
randomness into the degree value-ordering heuristic without adversely affecting its performance
\emph{in aggregate}.

The right-hand plot of \cref{figure:value-ordering-heuristics} gives a detailed comparison of the
degree and biased heuristics. It shows that despite the aggregate performance being similar, on a
case by case basis, the two heuristics can make a large difference to the performance of individual
satisfiable instances. This suggests that although degree is a good heuristic, we should perhaps not
commit heavily to a vertex of highest degree, but also consider vertices of the same or similar
degree.

@@

In \cref{figure:old-vs-new} we show the effects of adding restarts and nogood recording to the
algorithm. With restarts and nogood recording (random search with restarts, RSR), the random
value-ordering heuristic comfortably beats the degree strategy with depth-first search. In other
words, although having a good value-ordering heuristic is beneficial, introducing randomness into
the search is better, if it is done alongside a mechanism to avoid heavy commitment to any
particular random choice. However, the biased heuristic together
with restarts (solution-biased search, SBS) is better still---that is, if we are introducing
restarts, then it is better to add a small amount of randomness to a tailored heuristic than it is
to throw away the heuristic altogether.  Indeed, the original algorithm can solve ?? satisfiable
instances by ?? seconds, whilst the biased and random restarting algorithms require only ?? seconds
and ?? seconds respectively to solve the same number.

@@

?? No sharing, sharing with Luby restarts, sharing with constant restarts, sharing with triggered
restarts

?? Reproducibility, constant and triggered

?? How many more nodes on unsat instances

?? Scalability to MPI

@@

@@

Performance results from these two modified algorithms, using the same families of instances as in
the previous section, are shown in \cref{figure:mcs}. Although we have
moved from a decision problem to an optimisation problem, the same changes remain clearly
beneficial. For the $k{\downarrow}$ algorithm, the change has a minimal effect on many instances
(typically, where the $k = 0$ subproblem is unsatisfiable and hard, and the $k = 1$ subproblem is
satisfiable and easy), but gives large benefits on many more instances than it penalises: it is over
an order of magnitude better on nearly a hundred instances, whilst being an order of magnitude worse
on only seven.

@@

\section*{Acknowledgements}

The authors would like to thank Ruth Hoffmann, Christophe Lecoutre, Christine Solnon, and Craig
Reilly for their comments.

>>>>

\begin{abstract}
    The current state of the art in subgraph isomorphism solving involves using degree as a
    value-ordering heuristic to direct backtracking search. Such a search makes a heavy commitment
    to the first branching choice, which is often incorrect. To mitigate this, we introduce and
    evaluate a new approach, which we call ``solution-biased search''. By combining a
    slightly-random value-ordering heuristic, rapid restarts, and nogood recording, we achieve two
    orders of magnitude faster performance on satisfiable instances, whilst not affecting
    performance on unsatisfiable instances. This search can be parallelised in a very simple
    but effective way: across both satisfiable and unsatisfiable instances, we get a further speedup
    of over thirty from thirty six cores, or ninety five from ten distributed-memory hosts.  We also
    show that solution-biased search is also suitable for optimisation problems, by using it to
    improve two maximum common induced subgraph algorithms.
\end{abstract}

@@

The subgraph isomorphism problem is to decide whether a copy of a small ``pattern'' graph occurs
inside a larger ``target'' graph.  ?? Applications

Although the problem is \NP-complete, by combining design techniques from artificial intelligence
with careful algorithm engineering, modern subgraph isomorphism solvers can often produce exact
solutions quickly even on graphs with thousands of vertices. The current single strongest subgraph
isomorphism solver uses ``highest degree first'' as a value-ordering heuristic to direct a
constraint programming style search
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16,DBLP:journals/jair/McCreeshPST18}. This
heuristic is much better than branching randomly, but is still far from perfect. To offset mistakes
made by this heuristic, this paper proposes a new perspective on value-ordering: rather than
defining a search order, we use degree to direct what \emph{proportion} of the search effort should
be spent in each subproblem. By combining rapid restarts and nogood recording, and introducing a
small amount of randomness into the value-ordering heuristic, we make a state-of-the-art subgraph
algorithm perform two orders of magnitude better on a large number of satisfiable instances, whilst
performing worse only rarely on satisfiable instances, and never on unsatisfiable instances.  This
strategy is also effective in an optimisation setting, producing benefits in two maximum common
induced subgraph algorithms. Through this combination of techniques we have, for the first time, a
practical alternative to backtracking search for constraint programming style algorithms where we
have a strong value-ordering heuristic and where we care both about satisfiable and unsatisfiable
instances.

This new form of search can also be parallelised, with a \emph{much} simpler implementation than
conventional work-stealing. By running many threads with different random seeds but the same restart
schedule, and sharing nogoods only following restarts, we can achieve speedups of thirty one from a
thirty six core machine, or ninety five by using ten such machines with message passing.

@@

@@

\begin{description}
    \item[BVG(r), M4D(r), and Rand] are families of randomly generated graphs using different models (bounded
        degree, regular mesh, and uniform), where each pattern is a permuted random connected subgraph
        of the target (and so each instance is satisfiable) \cite{DBLP:journals/pami/CordellaFSV04}.
        These benchmark instances are widely used, but have unusual properties and so broad
        conclusions should not be drawn based solely upon behaviour on these instances
        \cite{DBLP:journals/jair/McCreeshPST18}.
    \item[SF] contains randomly generated scale-free graphs using a similar method
        \cite{DBLP:journals/constraints/ZampelliDS10}.
   \item[LV] consists of various kinds of graph gathered by Larrosa and Valiente
       \cite{DBLP:journals/mscs/LarrosaV02} from the Stanford Graph Database. We include both the
        50 small graphs, and the 50 larger graphs.
    \item[Phase] contains hard crafted instances that lie near the satisfiable / unsatisfiable phase
        transition \cite{DBLP:journals/jair/McCreeshPST18}.
    \item[PR] contains graphs generated from segmented images, corresponding to a computer vision
        problem \cite{DBLP:journals/pr/SolnonDHJ15}.
    \item[Images and Meshes] contain graphs representing 2D segmented images and 3D object models,
        again representing a computer vision problem \cite{DBLP:journals/cviu/DamiandSHJS11}.
\end{description}

@@

We begin with a set of baseline performance measurements. In the top
two plots \cref{figure:others} we show the cumulative number of instances solved over time for the
non-induced and induced problems respectively. We compare the Glasgow Subgraph Solver using
depth-first search (DFS) and with the modifications described in the remainder of this paper
(solution-biased search, SBS), the PathLAD variation of the LAD algorithm
\cite{DBLP:journals/ai/Solnon10,DBLP:conf/lion/KotthoffMS16}, VF2
\cite{DBLP:journals/pami/CordellaFSV04}, RI \cite{DBLP:journals/bmcbi/BonniciGPSF13}, and VF3
\cite{DBLP:conf/gbrpr/CarlettiFSV17} (which only supports the induced problem), in each case using
the original implementation provided by the algorithm's authors. For each algorithm, the $y$ value
gives the cumulative number of instances which (individually) can be solved in no more than $x$
milliseconds.  The vertical distance between two lines therefore shows how many more instances can
be solved by one solver than another, if every instance is run separately with the chosen $x$
timeout. The horizontal distance shows how many times longer the per-instance timeout would need to
be to allow the rightmost algorithm to succeed on $y$ out of the 14,621 instances (bearing in mind
that the two sets of $y$ instances could be different).  The plots show that our starting point
comfortably beats PathLAD, VF2, VF3 and RI, except for very low choices of timeout.

@@

Traditionally, value-ordering heuristics are designed to drive search towards the most promising
region of the search space \cite{DBLP:conf/ecai/Geelen92}, or the most constrained
\cite{DBLP:conf/aaai/GentMPW96}, or the region with the highest solution density
\cite{DBLP:journals/jair/PesantQZ12}. In subgraph isomorphism, this is done by selecting vertices
from highest degree to lowest \cite{DBLP:journals/jair/McCreeshPST18}. The left-hand plot of
\cref{figure:value-ordering-heuristics} demonstrates that this is indeed a good choice: the
\textbf{Degree} heuristic's line shows much better performance on satisfiable instances than the
\textbf{Random} (branch randomly) or \textbf{Anti} (branch from lowest degree to highest) heuristic
lines. Meanwhile, on unsatisfiable instances, the value-ordering heuristic has no effect on
performance, because a complete search must be performed.

But what happens if our value-ordering heuristic has to choose between mapping a pattern vertex to,
for example, three target vertices of degree ten, two vertices of degree nine, or five of degree
two? When driving conventional backtracking search, the degree heuristic would pick one of the
vertices of degree ten, and we would commit all of our search effort to the exponentially large
search tree underneath it, not considering any other choice until this tree has been fully explored
and eliminated.  We will show that this is not a wise choice, and that instead, we should
\emph{commit equal search effort} to each of the three subproblems found by mapping to vertices of
degree ten.  And similarly, should we be certain that vertex of degree ten is so much better than a
vertex of degree nine that we should commit no effort to degree nine vertices until all the degree
ten subproblems have been explored? Or might it be better to commit, say, twice as much effort to
each degree ten subproblems as to each degree nine subproblems, and only a very small amount of
effort to the degree two subproblems?  To test this hypothesis, we will now introduce a new
alternative to backtracking search, which we call \emph{solution-biased search}.  This search is
made up of three components: a new slightly-random value-ordering heuristic, rapid restarts, and
nogood recording. The aim is to perform a complete search, but spending proportionally more time in
parts of the search tree that are preferred by the value-ordering heuristic.

@@

Firstly, we define a new \textbf{Biased} value-ordering heuristic, as follows. When branching, we
select a vertex $v'$ from the chosen domain $D_v$ with probability \[ p(v') =
\frac{2^{\deg(v')}}{\sum_{w \in D_v}{2^{\deg(w)}}} \text{.} \] This heuristic is now equally likely
to pick between vertices of equal degree, is twice as likely to pick a vertex of degree $d$ as one
of degree $d - 1$, and is over a thousand times more likely to pick a vertex of degree $d$ than
degree $d - 10$.

\Cref{figure:value-ordering-heuristics} confirms that this heuristic, when used with backtracking
search, will solve close to the same number of instances as the degree heuristic would for any given
choice of timeout. In other words, we can introduce an element of randomness into the degree
value-ordering heuristic without adversely affecting its performance \emph{in aggregate}.  The
right-hand plot gives a detailed comparison. It shows that despite the aggregate performance being
similar, on a case by case basis, the two heuristics can make a large difference to the performance
for individual satisfiable instances. This justifies our belief that although degree is a good
heuristic, we should perhaps not commit heavily to a vertex of highest degree, but also consider
vertices of the same or similar degree.

@@

In \cref{figure:old-vs-new} we show the effects of adding restarts and nogood recording to the
algorithm. With restarts and nogood recording (random search with restarts, RSR), the random
value-ordering heuristic comfortably beats the degree strategy with depth-first search. In other
words, although having a good value-ordering heuristic is beneficial, introducing randomness into
the search is better, if it is done alongside a mechanism to avoid heavy commitment to any
particular random choice. However, the biased heuristic together
with restarts (solution-biased search, SBS) is better still---that is, if we are introducing
restarts, then it is better to add a small amount of randomness to a tailored heuristic than it is
to throw away the heuristic altogether.  Indeed, the original algorithm can solve 1983 satisfiable
instances by 909 seconds, whilst the biased and random restarting algorithms require only 12 seconds
and 35 seconds respectively to solve the same number.

@@

@@

Exploiting multiple cores to speed up constraint programming solvers remains an active area of
research, with no universally perfect solution being available. Four of the more common approaches
are based upon decompositions
\cite{DBLP:journals/corr/abs-1008-4328,DBLP:journals/jair/MalapertRR16}, work stealing
\cite{DBLP:conf/cp/MichelSH07,DBLP:conf/cp/ChuSS09,DBLP:conf/cp/McCreeshP15,DBLP:conf/cpaior/HoffmannMNPRS018},
parallel discrepancy searches \cite{DBLP:conf/cp/MoisanGQ13,DBLP:conf/cpaior/MoisanQG14}, and
algorithm portfolios \cite{DBLP:conf/lion/LindauerHH15}. Decomposition approaches are unsuitable for
decision problems, or problems where we have good value-ordering heuristics, because the
decomposition interferes strongly with the shape of the search tree
\cite{DBLP:journals/jair/MalapertRR16}. Work stealing, traditionally, also interferes with
value-ordering \cite{DBLP:journals/topc/McCreeshP15}, although specially designed exceptions exist
\cite{DBLP:conf/cp/ChuSS09,DBLP:conf/cp/McCreeshP15,DBLP:conf/cpaior/HoffmannMNPRS018}. However,
these have very complicated implementations. Parallel discrepancy searches are aware of
value-ordering heuristics, but have other limitations: they struggles on search trees with heavy
filtering, and rely upon inner search tree nodes being orders of magnitude less expensive to process
than leaf nodes. Portfolios, meanwhile, typically rely upon running multiple models or heuristics
simultaneously, and selecting whichever finishes first, whereas here we have a known good model and
set of heuristics.

\subsection{Shared Memory Parallelism}

Solution-biased search allows for a much simpler parallel implementation, which we now evaluate.  We
create a number of threads, and give each thread its own random seed, but otherwise each thread
performs the same sequential search. Then, whenever a thread restarts, it uses a barrier to wait for
every other thread to also restart. Each thread's new nogoods from that restart can then be
gathered, and then each thread resumes, with a larger set of nogoods than would have been generated
in a sequential run. Finally, whenever any single thread terminates, either due to having found a
solution or proved unsatisfiability, then every other thread may immediately terminate.

A number of factors combine to make this technique feasible:

\begin{itemize}
    \item Each thread will be run with the same restart schedule, and so will spend approximately
        the same amount of time between restarts. Because the only synchronisation between threads
        is at restarts, we expect threads to be kept busy doing useful work.  (This is in contrast
        to an alternative method for paralellising restarting search \cite{DBLP:conf/aaai/CireKS14},
        which packs together successive sequence values to produce a balanced workload.)
    \item Sequentially, on non-trivial instances the algorithm will restart very many times (many
        tens of thousands, for instances that reach the thousand second timeout).
    \item Because the search trees we explore are exponentially large, the randomness in the
        value-ordering heuristic is sufficient to ensure that most of the time, threads are
        exploring different parts of the search tree.
    \item The gathering of nogoods to describe the work done so far provides an alternative to
        requiring either a specific mechanism to allocate work, or expensive synchronisation between threads.
    \item If sometimes threads do happen to explore part of
        the same subproblems, this is not a problem: if the instance is satisfiable, either thread
        might find a solution first, and if the instance is unsatisfiable, we merely introduce some
        redundancy into the proof.\footnote{For solving a counting or enumeration problem,
        matters become slightly more complicated, but not devastatingly so.} The combination of
        rapid restarts and nogood recording is enough to ensure that this is only a small overhead.
\end{itemize}

\Cref{figure:parallel} shows how this scheme performs in practice. Sequentially, we can solve 14,357
instances within the thousand second timeout, with the last instance being solved at 939.0 seconds.
Using thirty six threads on machines with two eighteen core processors, we can solve 14,357
instances with a timeout of only 74.2 seconds, giving an aggregate speedup of 12.7.

Closer inspection of the results reveals that with this many threads, a considerable proportion of
the overall search time is spent with threads waiting at the barrier for synchronisation. This is
because the time taken to carry out search until 660 backtracks are encountered is only
\emph{roughly} a constant (in practice it usually varies by around a factor of two). Furthermore,
the Luby sequence includes occasional large multipliers, and if unsatisfiability is proved during
towards the end of one of these runs, each thread will end up duplicating a large amount of work.

Because we are using nogood recording, an alternative approach is possible. Rather than using the
Luby sequence for restarts, we could restart after either a constant number of backtracks, or simply
after a certain time interval has passed. \Cref{figure:parallel} also shows the effects of restarting
every 100ms. Sequentially, this approach is slightly better than using the Luby sequence, being able to
solve 14,370 instances, with the last at 996.4 seconds. With thirty six threads, solving this many
instances takes 31.8 seconds, giving an aggregate speedup of 31.4.

It is important to emphasise that there is no expectation of a linear speedup from this approach,
particularly for satisfiable instances. Due to the random nature of the value-ordering heuristic,
one of our extra threads may ``get lucky'' and find a solution much quicker than we would
sequentially, leading to a superlinear speedup. Conversely, our extra threads may contribute nothing to finding a
solution---or worse, due to new nogoods altering the choice made by the random branching heuristic,
we could even get an absolute slowdown. We can see both of these effects in the bottom left plot of
\cref{figure:parallel}: although we achieve roughly a linear speedup on unsatisfiable instances,
satisfiable instances show much greater variability. However, this approach does at least mirror our intuition of
allocating search effort in proportion to where the value-ordering heuristic believes it will be
most fruitful, and so we should not be too surprised that we see roughly a linear speedup on average
on harder instances.

Additionally, for easy instances, most of the algorithm's execution time is spent in a preprocessing
phase. We have not parallelised this, which is why our results are poor below the one second mark.

\subsection{Distributed Memory Parallelism}

To further test the scalability of this technique, we also implemented a distributed-parallelism
layer on top of the threaded layer, using MPI. In contrast to the huge difficulties of implementing
work-stealing in a distributed memory setting, this required only the addition of a simple ``all
gather'' operation to communicate nogoods, and a ``gather'' to collect and combine the results of
each host. Termination, meanwhile, was handled by posting an empty nogood (and so termination can
only occur on a restart).

\Cref{figure:parallel} also shows the results of these experiments, using five and ten hosts.
Because each host has two sockets, the five host results use ten MPI ranks, each with eighteen
threads, and the ten host results use twenty MPI ranks. Because the supercomputing service we use is
not designed for huge numbers of very short problems, we ran only instances whose sequential runtime
was at least one second, and for the sake of plotting results, treat skipped instances as taking one
second. Due to the job launcher used, it is also not possible to accurately measure ``total''
runtime including startup costs, and so instead we report the runtime of the rank zero
process---this figure is somewhat optimistic for very easy instances. With these caveats in mind, we
see that our results continue to improve as the host count increases, and we come very close to the
magic goal of an overall speedup of one hundred.

@@

Performance results from these two modified algorithms, using the same families of instances as in
the previous section, are shown in \cref{figure:mcs}. Although we have
moved from a decision problem to an optimisation problem, the same changes remain clearly
beneficial. For the $k{\downarrow}$ algorithm, the change has a minimal effect on many instances
(typically, where the $k = 0$ subproblem is unsatisfiable and hard, and the $k = 1$ subproblem is
satisfiable and easy), but gives large benefits on many more instances than it penalises: it is over
an order of magnitude better on over three hundred instances, whilst being an order of magnitude worse
on only seven.

@@

% \section*{Acknowledgements}
% 
% The authors would like to thank Ozgur Akgun, Chris Jefferson, Christophe Lecoutre, Christine
% Solnon, and Craig Reilly for their comments.
